---
title: "Thesis St. Andrews"
author: "Alvaro Fernandez de la Cigoña"
date: "2024-06-19"
output: html_document
---
Libraries

```{r, include=FALSE}
library(readr)
library(rstan)
library(dplyr)
library(tidyr)
library(brms)
library(rugarch)
library(mgcv)
library(caTools)
library(caret)
library(loo)
library(lubridate)
library(knitr)
```


Data Preparation
```{r}
# Loading data
blossoms <- read_csv("/Users/alvarofernandezdelacigonabarreiro/Documents/R St. Andrews/sakura_first_bloom_dates.csv")
head(blossoms)

#Dropping average column and notes
blossoms <- blossoms %>% select(-`30 Year Average 1981-2010`, -Notes)

#Reshaping format
reshaped_data <- blossoms %>%
  pivot_longer(cols = starts_with('1953'),  
names_to = "Year",
values_to = "Value")

head(reshaped_data)

# Dropping the 'Year' and 'Value' columns
reshaped_data <- reshaped_data %>% select(-Year, -Value)

#Dropping Wakkanai row because of NAs
reshaped_data <- reshaped_data %>% filter(!(`Site Name`%in% c("Wakkanai", "Rumoi")))

# Function to replace NA 
replace_na_with_previous <- function(row) {
  for (i in seq_along(row)) {
   if (is.na(row[i])) {
     row[i] <- row[i-1]
    }
  }
  return(row)
}

# Rellenar valores NA con el valor anterior en cada fila para las columnas de años
data_corrected <- reshaped_data %>%
  fill(`1954`:`2023`, .direction = "down")


# Viewing the corrected data
head(data_corrected)

# Converting date columns to numeric day of the year
data_numeric <- data_corrected %>%
  mutate(across(matches("^(19|20)\\d{2}$"), ~ yday(as.Date(.))))

# Creating a custom dataset to pass to brms
data_timeser <- data_numeric %>%
  pivot_longer(cols = matches("^(19|20)\\d{2}$"), names_to = "Year", values_to = "NumericDate") %>%
  mutate(Year = as.numeric(Year))

# Checking for duplicate year values within each group
duplicates <- data_timeser %>%
  group_by(`Site Name`, Year) %>%
  filter(n() > 1)

if (nrow(duplicates) > 0) {
  print("Duplicates found within groups:")
  print(duplicates)
}

data <- data_timeser
head(data)


data <- data %>%
  rename(SiteName = `Site Name`)


```
Visualization Time Series (hacerlo de solo un site mejor y ponerlo como ejemplo!!!)

```{R}
plot.ts(data$NumericDate)

acf(data_ar2$NumericDate)

```


Data handling  


```{R}
# Ensure the data is sorted by location and year
data <- data %>%
  arrange(SiteName,Year)

# Create lagged variables for AR(2) model
data_ar2 <- data %>%
  group_by(SiteName) %>%
  mutate(Value_lag1 = lag(NumericDate, 1),
         Value_lag2 = lag(NumericDate, 2)) %>%
  ungroup() %>%
  drop_na(c(NumericDate, Value_lag1, Value_lag2)) 

# Check for any remaining missing values
if (any(is.na(data_ar2))) {
  stop("There are still missing values in the data after creating lagged variables.")
}

# For short-term predictions 
# Leave the last 100 rows for validation
data_short_term <- data_ar2[1:(nrow(data_ar2) - 100), ]  

# For long-term predictions 
# Only the last 300 rows fortraining and validation
data_long_term <- data_ar2[(nrow(data_ar2) - 300):nrow(data_ar2), ] 

```


Data Partitioning (80/20)


```{R}
data_proportion <- 0.8
data_partitioning <- floor(data_proportion * nrow(data))


# Dividir conjunto de datos completo de forma aleatoria
set.seed(123)  
split <- sample.split(data$SiteName, SplitRatio = data_proportion)
training <- subset(data, split == TRUE)
test <- subset(data, split == FALSE)

# Dividir data_ar2 de forma similar
split_ar2 <- sample.split(data_ar2$SiteName, SplitRatio = data_proportion)
training_ar2 <- subset(data_ar2, split_ar2 == TRUE)
test_ar2 <- subset(data_ar2, split_ar2 == FALSE)

```


Cross Validation functions for different scenarios


```{R}
## Long-Term Prediction Cross-Validation
long_term_cv <- function(data, model_formula, forecast_horizon, family, chains = 2, iter = 2000, warmup = 1000) {
  training_long_term <- subset(data, Year <= max(Year) - forecast_horizon)
  validation_long_term <- subset(data, Year > max(training_long_term$Year))
}
  

  ## Short-Term Prediction Cross-Validation
short_term_cv <- function(data, model_formula, forecast_horizon, family, chains = 2, iter = 2000, warmup = 1000) {
  training_short_term <- subset(data, Year <= max(Year) - forecast_horizon)
  validation_short_term <- subset(data, Year > max(training_short_term$Year) & Year <= max(training_short_term$Year) + forecast_horizon)
}


```


Data partitioning


```{R}
#Long-Term scenario
long_term_partition <- function(data, forecast_horizon) {
  # Training set: All data up to the forecast_horizon before the latest year
  training_long_term <- subset(data, Year <= max(Year) - forecast_horizon)
  
  # Validation set: Data within the forecast_horizon (last 6 months)
  validation_long_term <- subset(data, Year > max(training_long_term$Year))
  
  return(list(training = training_long_term, validation = validation_long_term))
}


long_term_data <- long_term_partition(data_ar2, forecast_horizon = 0.5)  # 6 months = 0.5 years
training_long_term <- long_term_data$training
validation_long_term <- long_term_data$validation

#Short-Term Scenario
# Using the last 2 years for validation
short_term_partition <- function(data, recent_years) {
  # Training set: Using all data except the last 'recent_years' years
  training_short_term <- subset(data, Year <= max(Year) - recent_years)
  
  # Validation set: Using last 'recent_years' of data
  validation_short_term <- subset(data, Year > max(training_short_term$Year))
  
  return(list(training = training_short_term, validation = validation_short_term))
}

short_term_data <- short_term_partition(data_ar2, recent_years = 2)
training_short_term <- short_term_data$training
validation_short_term <- short_term_data$validation
```


Data Partitioning using K-fold
```{R}


# For short-term predictions 
# Leave the last 100 rows for validation
data_short_term <- data_ar2[1:(nrow(data_ar2) - 100), ]  

# For long-term predictions 
# Only the last 300 rows fortraining and validation
data_long_term <- data_ar2[(nrow(data_ar2) - 300):nrow(data_ar2), ] 

# Short-Term k-Fold Cross Validation
kst <- 5
num_observations_short <- nrow(data_short_term)
fold_size_short <- floor(num_observations_short / kst)

for (i in 1:kst) {
  train_end <- max(1, (i * fold_size_short))
  validation_start <- min(num_observations_short, train_end + 1)
  validation_end <- min(num_observations_short, (i + 1) * fold_size_short)
  
  training_set_kfold_short_term <- data_short_term[1:train_end, ]
  validation_set_kfold_short_term <- data_short_term[validation_start:validation_end, ]
}

# Long-Term k-Fold Cross Validation
k <- 5
num_observations_long <- nrow(data_long_term)
fold_size_long <- floor(num_observations_long / k)

for (i in 1:k) {
  train_end <- max(1, (i * fold_size_long))
  validation_start <- min(num_observations_long, train_end + 1)
  validation_end <- min(num_observations_long, (i + 1) * fold_size_long)
  
  training_set_kfold_long_term <- data_long_term[1:train_end, ]
  validation_set_kfold_long_term <- data_long_term[validation_start:validation_end, ]
}


```


AR(1) Model BRMS


```{r, include=FALSE}
# Fitting the AR(1) model
ar1_model <- brm(
  formula = NumericDate ~ Value_lag1 + (1 | SiteName),
  data = training_ar2,
  family = gaussian(),
  chains = 2, 
  iter = 2000,
  warmup = 500,
  control = list(adapt_delta = 0.95)
)

# Summarizing the model
summary(ar1_model)

# Predicting for 2033
predictions_AR1 <- posterior_predict(ar1_model, newdata = test_ar2, allow_new_levels =TRUE)

# Converting numeric date back to Date
predicted_date_2033_AR1 <- as.Date(round(mean(predictions_AR1)))

# Printting the predicted date for 2033
print(predicted_date_2033_AR1)



##Model evaluation (LOO-CV)

# Extracting log-likelihood values
log_lik_AR1 <- log_lik(ar1_model)

# Calculating relative effective sample sizes
r_eff_AR1 <- relative_eff(exp(log_lik_AR1), chain_id = rep(1:4, each = nrow(log_lik_AR1) / 4))

# Performing LOO-CV with r_eff
loo_result_AR1 <- loo(log_lik_AR1, r_eff = r_eff_AR1)

# Printing the LOO-CV results
print(loo_result_AR1)


##Model evaluation (RMS)

#Predictions on training data
predictions_ar1 <- posterior_predict(ar1_model, summary = FALSE)

#mean of predicted values for each point
predicted_means_ar1 <- rowMeans(predictions_ar1)


#Calculationg RMS
rmse_ar1 <- calculate_rmse_ar1(training_ar2$NumericDate, predicted_means_ar1)
print(paste("RMSE for the AR(1) model:", rmse_ar1))

```


AR(1) Model on different scenarios


```{R}
# Fitting the model
  AR1_long_term <- brm(
    formula = NumericDate ~ Value_lag1 + (1 | SiteName),
    data = training_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar1_long_term <- posterior_predict(AR1_long_term, newdata = validation_long_term)
  predicted_means_ar1_long_term <- rowMeans(predictions_ar1_long_term)
  
  rmse_ar1_long_term <- sqrt(mean((validation_long_term$NumericDate - predicted_means_ar1_long_term)^2))
  
  # LOO-CV
  loo_result_ar1_long_term <- loo(AR1_long_term)
  
  list(model = AR1_long_term, rmse = rmse_ar1_long_term, loo = loo_result_ar1_long_term)


  
  # Fitting the model
  AR1_short_term <- brm(
    formula = NumericDate ~ Value_lag1 + (1 | SiteName),
    data = training_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar1_short_term <- posterior_predict(AR1_short_term, newdata = validation_short_term)
  predicted_means_ar1_short_term <- rowMeans(predictions_ar1_short_term)
  
  rmse_ar1_short_term <- sqrt(mean((validation_short_term$NumericDate - predicted_means_ar1_short_term)^2))
  
  # LOO-CV
  loo_result_ar1_short_term <- loo(AR1_short_term)
  
  list(model = AR1_short_term, rmse = rmse_ar1_short_term, loo = loo_result_ar1_short_term)

```


AR(1) Model on different scenarios (K-fold)


```{R}
# Fitting the model
  AR1_long_term_kfold <- brm(
    formula = NumericDate ~ Value_lag1 + (1 | SiteName),
    data = training_set_kfold_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar1_long_term_kfold <- posterior_predict(AR1_long_term_kfold, newdata = validation_set_kfold_long_term, allow_new_levels = TRUE)
  predicted_means_ar1_long_term_kfold <- rowMeans(predictions_ar1_long_term_kfold)
  
  rmse_ar1_long_term_kfold <- sqrt(mean((validation_set_kfold_long_term$NumericDate - predicted_means_ar1_long_term_kfold)^2))
  
  # LOO-CV
  loo_result_ar1_long_term_kfold <- loo(AR1_long_term_kfold)
  
  list(model = AR1_long_term_kfold, rmse = rmse_ar1_long_term_kfold, loo = loo_result_ar1_long_term_kfold)


  
  # Fitting the model
  AR1_short_term_kfold <- brm(
    formula = NumericDate ~ Value_lag1 + (1 | SiteName),
    data = training_set_kfold_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar1_short_term_kfold <- posterior_predict(AR1_short_term_kfold, newdata = validation_set_kfold_short_term)
  predicted_means_ar1_short_term_kfold <- rowMeans(predictions_ar1_short_term_kfold)
  
  rmse_ar1_short_term_kfold <- sqrt(mean((validation_set_kfold_short_term$NumericDate - predicted_means_ar1_short_term_kfold)^2))
  
  # LOO-CV
  loo_result_ar1_short_term_kfold <- loo(AR1_short_term_kfold)
  
  list(model = AR1_short_term_kfold, rmse = rmse_ar1_short_term_kfold, loo = loo_result_ar1_short_term_kfold)

```


AR(2) Model BRMS


```{r, include=FALSE}
# Fitting the AR(2) model
ar2_model <- brm(
  formula = NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName),
  data = training_ar2,
  family = gaussian(),
  chains = 2, 
  iter = 2000,
  warmup = 500,
  control = list(adapt_delta = 0.95)
)

# Summarizing the model
summary(ar2_model)



# Predicting for 2033
predictions_AR2 <- posterior_predict(ar2_model, newdata = test_ar2, allow_new_levels =TRUE)

# Converting numeric date back to Date
predicted_date_2033_AR2 <- as.Date(round(mean(predictions_AR2)))

# Printting the predicted date for 2033
print(predicted_date_2033_AR2)



##Model evaluation (LOO-CV)

# Extracting log-likelihood values
log_lik_AR2 <- log_lik(ar2_model)

# Calculating relative effective sample sizes
r_eff_AR2 <- relative_eff(exp(log_lik_AR2), chain_id = rep(1:4, each = nrow(log_lik_AR2) / 4))

# Performing LOO-CV with r_eff
loo_result_AR2 <- loo(log_lik_AR2, r_eff = r_eff_AR2)

# Printing the LOO-CV results
print(loo_result_AR2)


##Model evaluation (RMS)

#Predictions on training data
predictions_ar2 <- posterior_predict(ar2_model, summary = FALSE)

#mean of predicted values for each point
predicted_means_ar2 <- rowMeans(predictions_ar2)


#Calculationg RMS
rmse_ar2 <- calculate_rmse_ar1(training_ar2$NumericDate, predicted_means_ar2)
print(paste("RMSE for the AR(2) model:", rmse_ar2))

```


AR(2) on different scenarios


```{R}

# Fitting the model
  AR2_long_term <- brm(
    formula = NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName),
    data = training_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar2_long_term <- posterior_predict(AR2_long_term, newdata = validation_long_term, allow_new_levels = TRUE)
  predicted_means_ar2_long_term <- rowMeans(predictions_ar2_long_term)
  
  rmse_ar2_long_term <- sqrt(mean((validation_long_term$NumericDate - predicted_means_ar2_long_term)^2))
  
  # LOO-CV
  loo_result_ar2_long_term <- loo(AR2_long_term)
  
  list(model = AR2_long_term, rmse = rmse_ar2_long_term, loo = loo_result_ar2_long_term)


  
  # Fitting the model
  AR2_short_term <- brm(
    formula = NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName),
    data = training_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar2_short_term <- posterior_predict(AR2_short_term, newdata = validation_short_term)
  predicted_means_ar2_short_term <- rowMeans(predictions_ar2_short_term)
  
  rmse_ar2_short_term <- sqrt(mean((validation_short_term$NumericDate - predicted_means_ar2_short_term)^2))
  
  # LOO-CV
  loo_result_ar2_short_term <- loo(AR2_short_term)
  
  list(model = AR2_short_term, rmse = rmse_ar2_short_term, loo = loo_result_ar2_short_term)


```


AR(2) on different scenarios (k-fold)


```{R}

# Fitting the model
  AR2_long_term_kfold <- brm(
    formula = NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName),
    data = training_set_kfold_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar2_long_term_kfold <- posterior_predict(AR2_long_term_kfold, newdata = validation_set_kfold_long_term, allow_new_levels = TRUE)
  predicted_means_ar2_long_term_kfold <- rowMeans(predictions_ar2_long_term_kfold)
  
  rmse_ar2_long_term_kfold <- sqrt(mean((validation_set_kfold_long_term$NumericDate - predicted_means_ar2_long_term_kfold)^2))
  
  # LOO-CV
  loo_result_ar2_long_term_kfold <- loo(AR2_long_term_kfold)
  
  list(model = AR2_long_term_kfold, rmse = rmse_ar2_long_term_kfold, loo = loo_result_ar2_long_term_kfold)


  
  # Fitting the model
  AR2_short_term_kfold <- brm(
    formula = NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName),
    data = training_set_kfold_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_ar2_short_term_kfold <- posterior_predict(AR2_short_term_kfold, newdata = validation_set_kfold_short_term, allow_new_levels = TRUE)
  predicted_means_ar2_short_term_kfold <- rowMeans(predictions_ar2_short_term_kfold)
  
  rmse_ar2_short_term_kfold <- sqrt(mean((validation_set_kfold_short_term$NumericDate - predicted_means_ar2_short_term_kfold)^2))
  
  # LOO-CV
  loo_result_ar2_short_term_kfold <- loo(AR2_short_term_kfold)
  
  list(model = AR2_short_term_kfold, rmse = rmse_ar2_short_term_kfold, loo = loo_result_ar2_short_term_kfold)


```


Generalized Autoregressive Conditional Heteroskedasticity (GARCH)


```{R, include= FALSE}

heteroskedastic_model <- brm(
  bf(NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName), sigma ~ s(Year, bs = "gp")),
  data = training_ar2,
  family = gaussian(),
  chains = 2,
  iter = 2000,
  warmup = 500,
  control = list(adapt_delta = 0.95)
)

# Summarizing the model
summary(heteroskedastic_model)



# Predicting for 2033
predictions_GARCHE <- posterior_predict(heteroskedastic_model, newdata = test_ar2, allow_new_levels = TRUE)

# Extracting the mean prediction
predicted_day_of_year_GARCHE <- round(mean(predictions_GARCHE))

# Converting Julian day back to Date
predicted_date_2033_GARCHE <- as.Date(paste0("2033-01-01")) + (predicted_day_of_year_GARCHE - 1)

# Printing the predicted date for 2033
print(predicted_date_2033_GARCHE)


##Model evaluation LOO-CV
# Extracting log-likelihood values
log_lik_GARCH <- log_lik(heteroskedastic_model)

# Calculating relative effective sample sizes
r_eff_GARCH <- relative_eff(exp(log_lik_GARCH), chain_id = rep(1:4, each = nrow(log_lik_GARCH) / 4))

loo_result_GARCH <- loo(log_lik_GARCH, r_eff = r_eff_GARCH)
print(loo_result_GARCH)

##Model evaluation (RMS)

#Predictions on training data
predictions_GARCH <- posterior_predict(heteroskedastic_model, summary = FALSE)

#mean of predicted values for each point
predicted_means_GARCH <- rowMeans(predictions_GARCH)

#Calculationg RMS
rmse_GARCH <- calculate_rmse_ar1(training_ar2$NumericDate, predicted_means_GARCH)
print(paste("RMSE for the GARCH model:", rmse_GARCH))

```


GARCH for different scenarios


```{R}

# Fitting the model
  GARCH_long_term <- brm(
    bf(NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName), sigma ~ s(Year, bs = "gp")),
    data = training_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GARCH_long_term <- posterior_predict(GARCH_long_term, newdata = validation_long_term, allow_new_levels = TRUE)
  predicted_means_GARCH_long_term <- rowMeans(predictions_GARCH_long_term)
  
  rmse_GARCH_long_term <- sqrt(mean((validation_long_term$NumericDate - predicted_means_GARCH_long_term)^2))
  
  # LOO-CV
  loo_result_GARCH_long_term <- loo(GARCH_long_term)
  
  list(model = GARCH_long_term, rmse = rmse_GARCH_long_term, loo = loo_result_GARCH_long_term)


  
  # Fitting the model
  GARCH_short_term <- brm(
    bf(NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName), sigma ~ s(Year, bs = "gp")),
    data = training_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GARCH_short_term <- posterior_predict(GARCH_short_term, newdata = validation_short_term, allow_new_levels = TRUE)
  predicted_means_GARCH_short_term <- rowMeans(predictions_GARCH_short_term)
  
  rmse_GARCH_short_term <- sqrt(mean((validation_short_term$NumericDate - predicted_means_GARCH_short_term)^2))
  
  # LOO-CV
  loo_result_GARCH_short_term <- loo(GARCH_short_term)
  
  list(model = GARCH_short_term, rmse = rmse_GARCH_short_term, loo = loo_result_GARCH_short_term)



```


GARCH for different scenarios (k-fold)


```{R}

# Fitting the model
  GARCH_long_term_kfold <- brm(
    bf(NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName), sigma ~ s(Year, bs = "gp")),
    data = training_set_kfold_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GARCH_long_term_kfold <- posterior_predict(GARCH_long_term_kfold, newdata = validation_set_kfold_long_term, allow_new_levels = TRUE)
  predicted_means_GARCH_long_term_kfold <- rowMeans(predictions_GARCH_long_term_kfold)
  
  rmse_GARCH_long_term_kfold <- sqrt(mean((validation_set_kfold_long_term$NumericDate - predicted_means_GARCH_long_term_kfold)^2))
  
  # LOO-CV
  loo_result_GARCH_long_term_kfold <- loo(GARCH_long_term_kfold)
  
  list(model = GARCH_long_term_kfold, rmse = rmse_GARCH_long_term_kfold, loo = loo_result_GARCH_long_term_kfold)


  
  # Fitting the model
  GARCH_short_term_kfold <- brm(
    bf(NumericDate ~ Value_lag1 + Value_lag2 + (1 | SiteName), sigma ~ s(Year, bs = "gp")),
    data = training_set_kfold_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GARCH_short_term_kfold <- posterior_predict(GARCH_short_term_kfold, newdata = validation_set_kfold_short_term, allow_new_levels = TRUE)
  predicted_means_GARCH_short_term_kfold <- rowMeans(predictions_GARCH_short_term_kfold)
  
  rmse_GARCH_short_term_kfold <- sqrt(mean((validation_set_kfold_short_term$NumericDate - predicted_means_GARCH_short_term_kfold)^2))
  
  # LOO-CV
  loo_result_GARCH_short_term_kfold <- loo(GARCH_short_term_kfold)
  
  list(model = GARCH_short_term_kfold, rmse = rmse_GARCH_short_term_kfold, loo = loo_result_GARCH_short_term_kfold)



```


GAM

```{R}
#Model formula using a spline for Year
formula_gam <- bf(
  NumericDate ~ s(Year, bs = "tp") + Value_lag1 + Value_lag2 + (1 | SiteName),
  sigma ~ s(Year, bs = "tp")  
)

# Fitting the model
gam_model <- brm(
  formula = formula_gam,
  data = training_ar2,
  family = gaussian(),
  chains = 2,      
  cores = 4,       
  iter = 2000,    
  warmup = 500    
)

# Summarizing the model
summary(gam_model)


# Predicting for 2033
predicted_day_of_year_GAM <- predict(gam_model, newdata = test_ar2)

# Converting Julian day back to Date
predicted_date_2033_GAM <- as.Date(paste0("2033-01-01")) + (predicted_day_of_year_GAM - 1)

# Printing the predicted date for 2033
print(predicted_date_2033_GAM)


##Model evaluation LOO-CV
loo_result_GAM <- loo(gam_model)
print(loo_result_GAM)

##Model evaluation (RMS)

#Predictions on training data
predictions_GAM <- posterior_predict(gam_model, summary = FALSE, allow_new_levels = TRUE)

#mean of predicted values for each point
predicted_means_GAM <- rowMeans(predictions_GAM)


#Calculationg RMS
rmse_GAM <- calculate_rmse_ar1(training_ar2$NumericDate, predicted_means_GAM)
print(paste("RMSE for the GAM model:", rmse_GAM))
```


GAM for different scenarios


```{R}
# Fitting the model
  GAM_long_term <- brm(
    formula = formula_gam,
    data = training_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GAM_long_term <- posterior_predict(GAM_long_term, newdata = validation_long_term, allow_new_levels = TRUE)
  predicted_means_GAM_long_term <- rowMeans(predictions_GAM_long_term)
  
  rmse_GAM_long_term <- sqrt(mean((validation_long_term$NumericDate - predicted_means_GAM_long_term)^2))
  
  # LOO-CV
  loo_result_GAM_long_term <- loo(GAM_long_term)
  
  list(model = GAM_long_term, rmse = rmse_GAM_long_term, loo = loo_result_GAM_long_term)


  
  # Fitting the model
  GAM_short_term <- brm(
    formula = formula_gam,
    data = training_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GAM_short_term <- posterior_predict(GAM_short_term, newdata = validation_short_term, allow_new_levels = TRUE)
  predicted_means_GAM_short_term <- rowMeans(predictions_GAM_short_term)
  
  rmse_GAM_short_term <- sqrt(mean((validation_short_term$NumericDate - predicted_means_GAM_short_term)^2))
  
  # LOO-CV
  loo_result_GAM_short_term <- loo(GAM_short_term)
  
  list(model = GAM_short_term, rmse = rmse_GAM_short_term, loo = loo_result_GAM_short_term)



```


GAM for different scenarios (k-fold)


```{R}
# Fitting the model
  GAM_long_term_kfold <- brm(
    formula = formula_gam,
    data = training_set_kfold_long_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GAM_long_term_kfold <- posterior_predict(GAM_long_term_kfold, newdata = validation_set_kfold_long_term, allow_new_levels = TRUE)
  predicted_means_GAM_long_term_kfold <- rowMeans(predictions_GAM_long_term_kfold)
  
  rmse_GAM_long_term_kfold <- sqrt(mean((validation_set_kfold_long_term$NumericDate - predicted_means_GAM_long_term_kfold)^2))
  
  # LOO-CV
  loo_result_GAM_long_term_kfold <- loo(GAM_long_term_kfold)
  
  list(model = GAM_long_term_kfold, rmse = rmse_GAM_long_term_kfold, loo = loo_result_GAM_long_term_kfold)


  
  # Fitting the model
  GAM_short_term_kfold <- brm(
    formula = formula_gam,
    data = training_set_kfold_short_term,
    family = gaussian(),
    chains = 2, 
    iter = 2000,
    warmup = 500,
    control = list(adapt_delta = 0.95)
  )
  
  # Predicting and calculating RMSE
  predictions_GAM_short_term_kfold <- posterior_predict(GAM_short_term_kfold, newdata = validation_set_kfold_short_term, allow_new_levels = TRUE)
  predicted_means_GAM_short_term_kfold <- rowMeans(predictions_GAM_short_term_kfold)
  
  rmse_GAM_short_term_kfold <- sqrt(mean((validation_set_kfold_short_term$NumericDate - predicted_means_GAM_short_term_kfold)^2))
  
  # LOO-CV
  loo_result_GAM_short_term_kfold <- loo(GAM_short_term_kfold)
  
  list(model = GAM_short_term_kfold, rmse = rmse_GAM_short_term_kfold, loo = loo_result_GAM_short_term_kfold)

```


Model comparisons


```{R}

##LOOIC
#Extracting only the looic score
loo_result_AR1_looic <- loo_result_AR1$estimates["looic", "Estimate"]
loo_result_AR1_looic_short_term <- loo_result_ar1_short_term$estimates["looic", "Estimate"]
loo_result_AR1_looic_long_term <- loo_result_ar1_long_term$estimates["looic", "Estimate"]
loo_result_AR1_looic_long_term_kfold <- loo_result_ar1_long_term_kfold$estimates["looic", "Estimate"]
loo_result_AR1_looic_short_term_kfold <- loo_result_ar1_short_term_kfold$estimates["looic", "Estimate"]

loo_result_AR2_looic <- loo_result_AR2$estimates["looic", "Estimate"]
loo_result_AR2_looic_short_term <- loo_result_ar2_short_term$estimates["looic", "Estimate"]
loo_result_AR2_looic_long_term <- loo_result_ar2_long_term$estimates["looic", "Estimate"]
loo_result_AR2_looic_long_term_kfold <- loo_result_ar2_long_term_kfold$estimates["looic", "Estimate"]
loo_result_AR2_looic_short_term_kfold <- loo_result_ar2_short_term_kfold$estimates["looic", "Estimate"]

loo_result_GARCH_looic <- loo_result_GARCH$estimates["looic", "Estimate"]
loo_result_GARCH_looic_short_term <- loo_result_GARCH_short_term$estimates["looic", "Estimate"]
loo_result_GARCH_looic_long_term <- loo_result_GARCH_long_term$estimates["looic", "Estimate"]
loo_result_GARCH_looic_long_term_kfold <- loo_result_GARCH_long_term_kfold$estimates["looic", "Estimate"]
loo_result_GARCH_looic_short_term_kfold <- loo_result_GARCH_short_term_kfold$estimates["looic", "Estimate"]

loo_result_GAM_looic <- loo_result_GAM$estimates["looic", "Estimate"]
loo_result_GAM_looic_short_term <- loo_result_GAM_short_term$estimates["looic", "Estimate"]
loo_result_GAM_looic_long_term <- loo_result_GAM_long_term$estimates["looic", "Estimate"]
loo_result_GAM_looic_long_term_kfold <- loo_result_GAM_long_term_kfold$estimates["looic", "Estimate"]
loo_result_GAM_looic_short_term_kfold <- loo_result_GAM_short_term_kfold$estimates["looic", "Estimate"]

#Putting them together
loo_models <- c(
  AR1_model = loo_result_AR1_looic, AR1_short_term = loo_result_AR1_looic_short_term, AR1_long_term = loo_result_AR1_looic_long_term, AR1_short_term_kfold = loo_result_AR1_looic_short_term_kfold, AR1_long_term_kfold = loo_result_AR1_looic_long_term_kfold, 
  AR2_model = loo_result_AR2_looic, AR2_short_term = loo_result_AR2_looic_short_term, AR2_long_term = loo_result_AR2_looic_long_term, AR2_short_term_kfold = loo_result_AR2_looic_short_term_kfold, AR2_long_term_kfold = loo_result_AR2_looic_long_term_kfold, 
  GARCH = loo_result_GARCH_looic, GARCHm_short_term = loo_result_GARCH_looic_short_term, GARCHm_long_term = loo_result_GARCH_looic_long_term, GARCHm_short_term_kfold = loo_result_GARCH_looic_short_term_kfold, GARCHm_long_term_kfold = loo_result_GARCH_looic_long_term_kfold, 
  GAM_model = loo_result_GAM_looic, GAM_model_short_term = loo_result_GAM_looic_short_term, GAM_model_long_term = loo_result_GAM_looic_long_term, GAM_model_short_term_kfold = loo_result_GAM_looic_short_term_kfold, GAM_model_long_term_kfold = loo_result_GAM_looic_long_term_kfold)

#Selection of the lowest looic score
best_loo <- names(which.min(loo_models))
print(best_loo)

#RMSE
rmse_models <- c(
  AR1_model = rmse_ar1,
  AR1_short_term = rmse_ar1_short_term,
  AR1_long_term = rmse_ar1_long_term,
  AR1_short_term_kfold = rmse_ar1_short_term_kfold,
  AR1_long_term_kfold = rmse_ar1_long_term_kfold,
    
  AR2_model = rmse_ar2, 
  AR2_short_term = rmse_ar2_short_term,
  AR2_long_term = rmse_ar2_long_term,
  AR2_short_term_kfold = rmse_ar2_short_term_kfold,
  AR2_long_term_kfold = rmse_ar2_long_term_kfold,
    
  GARCH_model = rmse_GARCH,
  GARCHm_short_term = rmse_GARCH_short_term,
  GARCHm_long_term = rmse_GARCH_long_term,
  GARCHm_short_term_kfold = rmse_GARCH_short_term_kfold,
  GARCHm_long_term_kfold = rmse_GARCH_long_term_kfold,
    
  GAM_model = rmse_GAM,
  GAM_model_short_term = rmse_GAM_short_term,
  GAM_model_long_term = rmse_GAM_long_term,
  GAM_model_short_term_kfold = rmse_GAM_short_term_kfold,
  GAM_model_long_term_kfold = rmse_GAM_long_term_kfold)

# Selection of the model with the lowest RMSE score
best_rmse <- names(which.min(rmse_models))
print(best_rmse)

```



Table summing up results


```{R}

# Long-Term Prediction Results
long_term_results <- data.frame(
  Model = c("AR1", "AR1_long_term", "AR1_long_term_kfold",
            "AR2", "AR2_long_term", "AR2_long_term_kfold",
            "GARCH", "GARCH_long_term", "GARCH_long_term_kfold",
            "GAM", "GAM_long_term", "GAM_long_term_kfold"),
  Scenario = "Long-Term Prediction",
  LOOIC_Estimate = c(49717.9, 61990.3, 2260.1,
                     49737.9, 61990.4, 2251.3,
                     447606.3, 58937.7, 1022.0,
                     47427.7, 58794.5, 2143.5),
  LOOIC_SE = c(819.6, 893.8, 736.5, 
               821.0, 895.0, 85.9,
               915.5, 1019.5, 50.6,
               976.6, 1074.0, 48.1),
  
  RMSE = c(32.8668769934646, 25.25375, 11.0288527, 
           32.8679509811609, 25.43484, 10.654226,
           32.9490890745172, 24.01065, 6.121305,
           32.8661373152288, 23.04074, 4.99656411)
)

# Short-Term Prediction Results
short_term_results <- data.frame(
  Model = c("AR1", "AR1_short_term", "AR1_short_term_kfold",
            "AR2", "AR2_short_term", "AR2_short_term_kfold", 
            "GARCH", "GARCH_short_term", "GARCH_short_term_kfold", 
            "GAM", "GAM_short_term", "GAM_short_term_kfold"),
  Scenario = "Short-Term Prediction",
  LOOIC_Estimate = c(49717.9, 61159.2, 61972.6, 
                     49737.9, 61152.2, 3264.7,
                     47606.3, 58255.5, 58743.7,
                     47427.7, 57883.5, 58474.4),
  LOOIC_SE = c(819.6, 879.8, 898.7, 
               821.0, 880.8, 110.3,
               915.5, 1029.0, 1022.0,
               976.6, 1035.6, 1094.3),
  RMSE = c(32.8668769934646, 24.9667, 24.916429, 
           32.8679509811609, 25.09681, 24.831462,
           32.9490890745172, 24.21252, 22.9267711,
           32.8661373152288, 24.38659, 23.382039)
)

kable(long_term_results)
kable(short_term_results)

# Combining all results into a single table
summary_table <- rbind(long_term_results, short_term_results)
print(summary_table)
kable(summary_table, caption = "Summary of LOOIC and RMSE for Different Models and Scenarios")


write.csv(long_term_results, "long_term_results1.csv")
write.csv(short_term_results, "short_term_results1.csv")

```

