---
title: "Knowledge Disvoc. Project 1"
author: "Alvaro Fernande de la Cigoña"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(LiblineaR)
library(ggplot2)
library(dslabs)
library(caret)
library(LiblineaR)
library(Metrics)
library(rpart)
library(randomForest)
library(rpart.plot)
```

UPLOADING THE DATA

```{r}
data_2022 <- read.csv("/Users/alvarofernandezdelacigonabarreiro/Documents/R St. Andrews/Combined_Flights_med_2022.csv")
data_2021 <- read.csv("/Users/alvarofernandezdelacigonabarreiro/Documents/R St. Andrews/Combined_Flights_med_2021.csv")
data_2020 <- read.csv("/Users/alvarofernandezdelacigonabarreiro/Documents/R St. Andrews/Combined_Flights_med_2020.csv")
data_2019 <- read.csv("/Users/alvarofernandezdelacigonabarreiro/Documents/R St. Andrews/Combined_Flights_med_2019.csv")
data_2018 <- read.csv("/Users/alvarofernandezdelacigonabarreiro/Documents/R St. Andrews/Combined_Flights_med_2018.csv")

combined_data <- bind_rows(data_2018, data_2019, data_2020, data_2021, data_2022)

data_1 <- combined_data[, !names(combined_data) %in% c("Cancelled", "Diverted", "ArrDel15", "FlightDate", "Operated_or_Branded_Code_Share_Partners", "DOT_ID_Marketing_Airline", "IATA_Code_Marketing_Airline", "Flight_Number_Marketing_Airline", "Originally_Scheduled_Code_Share_Airline", "DOT_ID_Originally_Scheduled_Code_Share_Airline", "IATA_Code_Originally_Scheduled_Code_Share_Airlin", "Flight_Num_Originally_Scheduled_Code_Share_Airline", "Operating_Airline", "DOT_ID_Operating_Airline", "IATA_Code_Operating_Airline", "Tail_Number", "Flight_Number_Operating_Airline", "OriginAirportSeqID", "OriginCityMarketID", "OriginCityName", "Origin", "OriginState", "OriginStateFips", "OriginStateName", "OriginWac", "DestAirportSeqID", "DestCityMarketID", "Dest", "DestCityName", "DestState", "DestStateFips", "DestStateName", "DestWac", "DepDelay", "DepDelayMinutes", "DepDel15", "DepartureDelayGroups", "DepTimeBlk", "TaxiOut", "WheelsOff", "WheelsOn", "TaxiIn", "ArrDelay", "ArrDelayMinutes", "ArrDel15", "ArrivalDelayGroups", "ArrTimeBlk", "CancellationCode", "CRSElapsedTime", "ActualElapsedTime", "DistanceGroup", "CarrierDelay", "AirTime", "WeatherDelay", "NASDelay", "LateAircraftDelay", "FirstDepTime", "TotalAddGTime", "LongestAddGTime", "DivAirportLandings", "DivReachedDest", "DivActualElapsedTime", "DivArrDelay", "DivDistance", "Div1Airport", "Div1AirportID	", "Div1AirportSeqID", "Div1WheelsOn", "Div1TotalGTime", "Div1LongestGTime", "Div1WheelsOff", "Div1TailNum", "Div2Airport", "Div2AirportID", "Div2AirportSeqID", "Div2WheelsOn", "Div2TotalGTime", "Div2LongestGTime", "Div2WheelsOff", "Div2TailNum", "Div3Airport", "Div3AirportID", "Div3AirportSeqID", "Div3WheelsOn", "Div3TotalGTime", "Div3LongestGTime", "Div3WheelsOff", "Div3TailNum", "Div4Airport", "Div4AirportID", "Div4AirportSeqID", "Div4WheelsOn", "Div4TotalGTime", "Div4LongestGTime", "Div4WheelsOff", "Div4TailNum", "Div5Airport", "Div5AirportID", "Div5AirportSeqID", "Div5WheelsOn", "Div5TotalGTime", "Div5LongestGTime", "Div5WheelsOff", "Div5TailNum", "Duplicate", "DayOfWeek", "Quarter", "DayofMonth")]

```

SPLITTING THE DATA (train and test)
Also dividing this sets into predictors and y variable(disruption); additionally filling the missing values of the y variable with 0s

```{r}
split_train_test <- function(data, test_ratio){
  shuffled_indices <- sample(1:nrow(data))
  test_set_size <- as.integer(nrow(data)*test_ratio)
  test_indices <- shuffled_indices[1:test_set_size]
  train_indices <- shuffled_indices[(test_set_size+1):nrow(data)]
  sets_out <- list(train=data[train_indices,], test=data[test_indices,])
  return(sets_out)
}

train_test_sets <- split_train_test(data_1, 0.2)
training_set <- train_test_sets$train
test_set <- train_test_sets$test

#Divide predictors and predicted variables on train set
y_train <- training_set$Disruption
y_train <- as.data.frame(y_train)
x_train <- subset(training_set, select = -Disruption)
#fill missing values on y
y_train[is.na(y_train)] <- 0

#Divide predictors and predicted variables on test set
y_test <- test_set$Disruption
y_test <- as.data.frame(y_test)
x_test <- subset(test_set, select = -Disruption)
#fill missing values on y
y_test[is.na(y_test)] <- 0
```


DATA VISUALISATION

````{R}
training_set_vis <- cbind(x_train, y_train)

training_set_vis |> 
  filter(y_train==1) |>
  ggplot(aes(x=Year)) + 
           geom_histogram(bins=5)

training_set_vis |>
  group_by(Marketing_Airline_Network) |>
  summarise(prop_disrup = mean(y_train)) |>
  ggplot(aes(x=Marketing_Airline_Network, y=prop_disrup)) +geom_col()

training_set_vis |>
  group_by(Distance) |>
  summarise(prop_disrup = mean(y_train)) |>
  ggplot(aes(x=Distance, y=prop_disrup)) +geom_col()

training_set_vis |>
  group_by(Year) |>
  summarise(prop_disrup = mean(y_train)) |>
  ggplot(aes(x=Year, y=prop_disrup)) +geom_col(bin=5)

training_set_vis |>
  group_by(Month) |>
  summarise(prop_disrup = mean(y_train)) |>
  ggplot(aes(x=Month, y=prop_disrup)) + geom_col(bin = 12)
````

SCALING AND ENCODING TRAIN SET

```{R}
##Scaling numerical values TRAIN
data_numerical_train <- select(x_train, -Year, -Month, -Airline, -Marketing_Airline_Network, -OriginAirportID, -DestAirportID)
# Calculate medians.
data_num_medians_train <- data_numerical_train %>% summarise(across(everything(), \(x) median( x, na.rm=TRUE)))
# replace nas with calculate medians
data_num_def_train <- data_numerical_train %>% tidyr::replace_na(as.list(data_num_medians_train))
#standarize num data
data_num_def1_train <- data_num_def_train %>% mutate(across(everything(), scale))


##One-hot encoder for categorical TRAIN
one_hot_encoder_train <- dummyVars("~ Year + Month + Airline + Marketing_Airline_Network + OriginAirportID + DestAirportID", data=x_train)
categorical_encoded_train <- data.frame(predict(one_hot_encoder_train, newdata = x_train))
head(categorical_encoded_train)
data_cat_def_train <- categorical_encoded_train

#combining cat and num
training_data_x <- bind_cols(data_num_def1_train, data_cat_def_train)
````

SCALING AND ENCODING TEST SET

```{R}
##Scaling numerical values TEST
data_numerical_test <- select(x_test, -Year, -Month, -Airline, -Marketing_Airline_Network, -OriginAirportID, -DestAirportID)
# Calculate medians.
data_num_medians_test <- data_numerical_test %>% summarise(across(everything(), \(x) median( x, na.rm=TRUE)))
# replace nas with calculate medians
data_num_def_test <- data_numerical_test %>% tidyr::replace_na(as.list(data_num_medians_test))
#standarize num data
data_num_def1_test <- data_num_def_test %>% mutate(across(everything(), scale))


###one-hot encoder for categorical TEST
one_hot_encoder_test <- dummyVars("~ Year + Month + Airline + Marketing_Airline_Network + OriginAirportID + DestAirportID", data=x_test)
categorical_encoded_test <- data.frame(predict(one_hot_encoder_test, newdata = x_test))
head(categorical_encoded_test)
data_cat_def_test <- categorical_encoded_test

#combining cat and num
testing_data_x <- bind_cols(data_num_def1_test, data_cat_def_test)
````

SGD CLASSIFIER

```{R}
set.seed(1234)
sgd_classifier <- LiblineaR(data=training_data_x, target = y_train)
predictions <- predict(sgd_classifier, newx=matrix(testing_data_x[1,], nrow=1))

#measure accuracy on test set
predictions <- predict(sgd_classifier, newx=testing_data_x)
sum(predictions$predictions == y_test) / nrow(x_test)
summary(x_test)

#predictions
test_set_predictions <- predict(sgd_classifier, newx=testing_data_x)
test_set_predictions
test_set_preds <- as.numeric(test_set_predictions$predictions==TRUE)
test_set_preds
test_set_labs <- as.numeric(y_test==TRUE)
test_set_labs
# calculate precision
test_precision <- Metrics::precision(test_set_labs, test_set_preds)
test_precision

#test recall
test_recall <- Metrics::recall(test_set_labs, test_set_preds)
test_recall

#f1
test_f1 <- 2 * (test_precision * test_recall)/(test_precision + test_recall)
test_f1
````

PRECISSION VS RECALL CURVE

```{R}
#output includes predictions and scores
y_valid_output <- predict(sgd_classifier, newx=testing_data_x, decisionValues = TRUE)
#get just score
y_valid_score <- y_valid_output$decisionValues[,1]

get_precisions <- function(threshold, score, labels){
  predictions <- score > threshold
  TP <- sum((predictions == TRUE) & (labels == TRUE))
  FP <- sum((predictions == TRUE) & (labels == FALSE))
  return(TP/(TP+FP))
}

get_recalls <- function(threshold, score, labels){
  predictions <- score > threshold
  TP <- sum((predictions == TRUE) & (labels == TRUE))
  FN <- sum((predictions == FALSE) & (labels == TRUE))
  return(TP/(TP+FN))
}

# create a set of thresholds
thresholds <- seq(min(y_valid_score)+0.01, max(y_valid_score)-0.01, length.out=1000)
# get precision and recall
precisions <- sapply(thresholds, get_precisions, score=y_valid_score, labels=y_test)
recalls <- sapply(thresholds, get_recalls, score=y_valid_score, labels=y_test)
ggplot() + geom_line(aes(x=recalls, y=precisions), color='blue') + 
  xlab("Recall") + ylab("Precision") + theme_bw(base_size=18)
````

ROC CURVE SGD

```{R}
get_FPRs <- function(threshold, score, labels){
  predictions <- score > threshold
  FP <- sum((predictions == TRUE) & (labels == FALSE))
  TN <- sum((predictions == FALSE) & (labels == FALSE))
  FPR <- FP / (FP+TN)
  return(FPR)
}

FPRs <- sapply(thresholds, get_FPRs, score=y_valid_score, labels = y_test)

ggplot() + geom_line(aes(x=FPRs, y=recalls), color='blue', linewidth=2) + 
  xlab("FPR") + ylab("TPR") + 
  theme_bw(base_size=18) + 
  geom_abline(linetype=4, lionewidth=2)
````

DECISSION TREE AND RANDOM FOREST

```{R}
set.seed(1234)
# set thresholds - decision tree and random forest output probabilities
prob_thresholds <- seq(0, 1, length.out=1000)

# fit decision tree
tree_classifier <- rpart(y_train$y_train~., training_data_x)
tree_predictions <- predict(tree_classifier, newdata=testing_data_x)
summary(tree_predictions)
tree_FPRs <- sapply(prob_thresholds, get_FPRs, score=tree_predictions, labels=y_test)
tree_FPRs
tree_recalls <- sapply(prob_thresholds, get_recalls, score=tree_predictions, labels=y_test)
tree_recalls

# fit random forest
forest_classifier <- randomForest(x=training_data_x, y=y_train$y_train, ntree=2, mtry=20)
forest_predictions <- predict(forest_classifier, newdata=testing_data_x, type="response")
forest_FPRs <- sapply(prob_thresholds, get_FPRs, score=forest_predictions, labels=y_test)
forest_FPRs
forest_recalls <- sapply(prob_thresholds, get_recalls, score=forest_predictions, labels=y_test)
forest_recalls
````

FINAL ROC CURVE

```{R}
cols <- c("SVM" = "blue", "tree" = "red", "forest"="purple")
ggplot() + geom_line(aes(x=FPRs, y=recalls, color="SVM"), linewidth=1.5) + 
  geom_line(aes(x=tree_FPRs, y=tree_recalls, color="tree"), linetype=2, linewidth=1.5) +
  geom_line(aes(x=forest_FPRs, y=forest_recalls, color="forest"), linetype=5, linewidth=1.5) + 
  xlab("FPR") + ylab("TPR") + theme_bw(base_size=18) + 
  geom_abline(linetype=4) + scale_color_manual(values = cols)
````

TUNING BEST MODEL (decision tree)

```{R}
test_x_y_union <- bind_cols(testing_data_x, y_test)
train_x_y_union <- bind_cols(training_data_x, y_train)
control <- rpart.control(minsplit = 10, minbucket = round(10/3), cp = 0,
                         maxcompete = 4)
fit <- rpart(y_train~., data = train_x_y_union, method = 'class', control = control)
````

INFO DECISION TREE TUNED

```{R}
tuned_model <- predict(fit, newdata = test_x_y_union)
summary(tuned_model)
tree_predictions_fit <- predict(fit, newdata=testing_data_x, type="prob")[,2]
tree_FPRs_fit <- sapply(prob_thresholds, get_FPRs, score=tree_predictions_fit, labels=y_test)
tree_recalls_fit <- sapply(prob_thresholds, get_recalls, score=tree_predictions_fit, labels=y_test)
ggplot() + geom_line(aes(x=tree_FPRs_fit, y=tree_recalls_fit), color='blue', linewidth=2) + 
  xlab("FPR") + ylab("TPR") + 
  theme_bw(base_size=18) + 
  geom_abline(linetype=4, lionewidth=2)

# Confusion Matrix
val<-mean(tree_predictions_fit)
resp<-ifelse(tree_predictions_fit>val,1,0)
my_conf <- table(as.numeric(resp), y_test$y_test)
my_conf
accuracy_Test <- sum(diag(my_conf))/sum(my_conf)
print(paste('Accuracy for test', accuracy_Test))
````

